---
title: "Exercise 8"
author: "Tobias Raidl, 11717659"
date: "2023-12-12"
output:
  pdf_document:
    toc: true
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)
```

```{r}
library(ggplot2movies)
data(movies)
df = movies[movies$Short == 1 & movies$year >= 2000, ]
```

# 1
Multiple correlation analysis: Compute the multiple correlation between the
variable rating and the matrix consisting of the variables year, length, budget,
votes. Delete observations containing missing values. It might be advisable to
transform “budget” and “votes”.
```{r}
library(dplyr)
df = na.omit(df)
x = df$rating
Y = select(df, c(year, length, budget, votes))
Y

hist(Y$budget)
Y$budget = log(Y$budget)
hist(Y$budget)
hist(Y$votes)
Y$votes = log(Y$votes)
hist(Y$votes)
```

## a
Compute the multiple correlation coefficient. How can you interpret the
coefficients of the linear predictor function?
```{r}
covmat_xY = cov(cbind(x, Y))
covmat_YY = covmat_xY[-1, -1]
cov_xY = covmat_xY[-1, 1]
str(t(cov_xY))
str(solve(covmat_YY))
str(cov_xY)
numerator = t(cov_xY) %*% solve(covmat_YY) %*% cov_xY
mult_cor_coef = sqrt(numerator/ var(x))
```

## b
 Is the multiple correlation coefficient significantly different from zero?
```{r}

```

## c
Use the function CCAgrid from the R package ccaPP – see help. Set the argument method="pearson" and compare the results with those from above.
Use method="spearman" and compare with the previous results. What is
the methodological difference?
```{r}
library(ccaPP)

CCAgrid(x, Y, method="pearson")

CCAgrid(x, Y, method="spearman")
```
The Pearson Correlation Coefficient assesses the linear relationship between variables, while the Spearman Correlation Coefficient evaluates the monotonic relationship. Spearman does not assume normally distributed data.

## d
Use the function permTest from the library(ccaPP). This function is performing a permutation test for uncorrelatedness, by permuting the observations of the first input. How and why does this work? What is the outcome?
Compare with the result in (b).
```{r}
permTest(x, Y)
```

# 2
Canonical correlation analysis: Compute the canonical correlation between
the matrices consisting of the variables year, length, budget, rating, votes and the
variables Action, Animation, Comedy, Drama, Documentary, Romance. Select
(transform) the observations according to the instructions at the beginning.

## a
 Use the function cancor() – see help. Center and scale the data (why?).
How strong is the linear relationship? How can you interpret the linear
combinations for the X and Y data?
```{r}
X2 = select(df, c(year, length, budget, rating, votes))
Y2 = select(df, c(Action, Animation, Comedy, Drama, Documentary, Romance))
X2_scaled = scale(X2)
Y2_scaled = scale(Y2)
cancor = cancor(X2_scaled, Y2_scaled)
```


