---
title: "Exercise 1"
author: "Tobias Raidl, 11717659"
date: "2023-10-13"
output: pdf_document
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)
```

### Setup Dataset
```{r}
library("pgmm")
data(wine)
wine
```
### 1.
By using a parallel coordinate system, colored according to each observations TYPE, you can estimate the explanatory power of each variable according to its TYPE. If the colored lines are clustered well it corresponds to a high explanatory power. This only holds if TYPE is perceived as dependent variable.
```{r}
library(MASS)
parcoord(wine, col=wine$Type)
```

### 2.
The two variables with the highest difference in mean when aggregated by TYPE are Proline and 2-3 Butanediol. You could compare the mean 
```{r}
library(dplyr)
grouped_means = wine %>%
  group_by(Type) %>%
  summarise_all(list(mean))

#rownames(grouped_means) = grouped_means$Type
grouped_means

grouped_means_diff = wine %>%
  summarise_all(list(function(x) diff(range(x)))) %>%
  select(-Type)
grouped_means_diff
```

### 3.
```{r}
grouped_variances = wine %>%
  group_by(Type) %>%
  summarise_all(list(var))
grouped_variances

grouped_variances_diff = wine %>%
  summarise_all(list(function(x) diff(range(x)))) %>%
  select(-Type)
grouped_variances_diff
```

### 4.
For the 2 variables with highest mean difference
```{r}
highest_diff_variable = colnames(grouped_means_diff)[apply(grouped_means_diff,1,which.max)]
temp = grouped_means_diff %>%
  select(-highest_diff_variable)

second_highest_diff_variable = colnames(temp)[apply(temp,1,which.max)]

print(paste("Variables with highest mean difference grouped by Type are: ", c(highest_diff_variable, second_highest_diff_variable)))

max_diff_vars = select(wine, contains(highest_diff_variable), contains(second_highest_diff_variable), Type)

parcoord(max_diff_vars, col=wine$Type)
plot(max_diff_vars, col=wine$Type)
```
For the 2 variables with highest variance difference
```{r}
max(grouped_variances_diff)

highest_diff_variable = colnames(grouped_variances_diff)[apply(grouped_variances_diff,1,which.max)]
temp = grouped_variances_diff %>%
  select(-highest_diff_variable)
second_highest_diff_variable = colnames(temp)[apply(temp,1,which.max)]

max_diff_vars = select(wine, contains(highest_diff_variable), contains(second_highest_diff_variable), Type)

parcoord(max_diff_vars, col=wine$Type)
plot(max_diff_vars, col=wine$Type)
```

### 5.
```{r}
cor_mat = cor(wine)
print(cor_mat)
heatmap(cor_mat)
```

### 6.
```{r}
eigen_error = function(k){
  temp = eigen(cor_mat)
  eigenvalues = temp$values[1:k]
  eigenvectors = temp$vectors[,1:k]
  eigenvalue_matrix = diag(eigenvalues)
  eigenvector_matrix = rbind(eigenvectors)
  recycled_cor_mat = eigenvector_matrix %*% eigenvalue_matrix %*% t(eigenvector_matrix)
  error = mean((cor_mat-recycled_cor_mat)^2)
  return(error)
}
error = eigen_error(5)
v = 2:ncol(wine)
v = unlist(lapply(v, FUN=eigen_error))
plot(v)
```

### 7
```{r}
library(ggplot2)

wine_maha = wine %>%
  mutate(maha = mahalanobis(wine, center=colMeans(wine), cov(wine))) %>%
  mutate(maha_grouped = case_when(
    Type == 1 ~ mahalanobis(wine, center=unlist(grouped_means[1,]), cov(wine)),
    Type == 2 ~ mahalanobis(wine, center=unlist(grouped_means[2,]), cov(wine)),
    Type == 3 ~ mahalanobis(wine, center=unlist(grouped_means[3,]), cov(wine))
    ))
print(wine_maha)

ggplot() + 
  geom_point(data = wine_maha, mapping = aes(x=Type, y=maha), color="blue") + 
  geom_point(data = wine_maha, mapping = aes(x=Type, y=maha_grouped), color="red")

```