---
title: "Exercise 6"
author: "Tobias Raidl, 11717659"
date: "2023-11-30"
output:
  pdf_document:
    toc: true
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)
```

# 1 Principal Component Analysis
```{r}
load("darwinM.RData")
df = darwinM
```

## a
Perform classical PCA on the whole data (except “class”), and try to explain
based on the biplot the main differences between the patient and the healthy
group.
```{r}
library(ggplot2)
library(ggfortify)
set.seed(11717659)
pca = princomp(~.-class, df, cor=TRUE)
autoplot(pca, data=df, colour="class", loadings = TRUE, loadings.colour = 'red',
         loadings.label = TRUE, loadings.label.size = 3)
```
The pressure mean is higher for healthy people, while the pressure variance is higher patients. Both air and paper time are higher for patients while mean speed in air and on paper both tend to be higher for healthy people.

## b
For the following tasks, use the function PcaHubert() from the package rrcov,
which performs a robust PCA.
Apply PCA separately on the healthy and on the patient group, and show the
PCA diagnostics plots with orthogonal and score distances. What could be the
reason for the clear outlyingness of some observations?
```{r}
library(rrcov)
library(dplyr)
pca_hubert = PcaHubert(~.-class, df, k=2, cor=TRUE)
df_h = select(df[df$class=="H", ], -class)
df_p = select(df[df$class=="P", ], -class)

pca_hubert_h = PcaHubert(~., df_h, k=2, scale=TRUE)
pca_hubert_p = PcaHubert(~., df_p, k=2, scale=TRUE)

par(mfrow = c(1, 2))
plot(pca_hubert_h, main="healthy")
plot(pca_hubert_p, main="patients")
```

## c
Present both biplots and try to identify main differences in the data structure of
both groups.
```{r}
biplot(pca_hubert_h, cex=0.7)
biplot(pca_hubert_p, cex=0.7)
```
For the patient group the loadings point in directions within circa 180 degrees, whereas the loadings from the healthy group have a higher variance in direction.

Ask Prof. Filzmoser: Cannot see much here. Any better way for visualization?

## d
Take the PCA result for the healthy group, and project the patient data into the
plane of the first 2 healthy group PCs. This allows for a better comparison of
the main differences in the data structure of the two data sets. What are these
main differences?
```{r}
biplot(pca_hubert_h, cex=0.7)
points(pca_hubert_p$scores, col="green", cex=0.7)
```
Sadly I cannot she anything in this plot.

Ask Prof. Filzmoser: Any better visualization method?

# 2
## a
Compute a factor analysis model for the complete data (without “class”) using
an appropriate number k of factors.
```{r}
fa <- factanal(scale(select(df, -class)),factors=5, scores="regression")
fa
```

## b
What is the maximum value of k you could consider?
k=8 is the largest k to consider. (formula in script)

## c
What are the differences to PCA (loadings, scores)? What are the uniquenesses?
Uniquenesses are the variances of the error terms. There is one uniqueness value per variable. The higher it is the less variance of it is explained by the factors.
 
 ## d
Show loadings and scores in a biplot and compare with the PCA biplot. How
can you interpret the first two factors?
The first factor focuses on all the horizontally displayed loading vectors like paper_time and pressure_mean. The second one on the mean_acc_on_paper and mean_jerk_on_paper. Variables whose loading vectors are rather short are hardly covered by the first 2 factors (max_y_extension, max_x_extension). They seem to be less relevant than others.
```{r}
autoplot(fa, data=df, colour="class", loadings = TRUE, loadings.colour = 'red',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca, data=df, colour="class", loadings = TRUE, loadings.colour = 'red',
         loadings.label = TRUE, loadings.label.size = 3)
```
